Approximation as Differentiation

### 2025-06-19 main/first-order.tex
Some clarification on lifting.

For a first-order language with a primitive type $\syn{nat}$ and primitive operations such as $+$, the
language would provide a model of $\syn{nat}$ and its operations in $\Fam(\LatGal)$. For example, we might
interpret $\syn{nat}$ to the object $(\Nat,n \mapsto \downset{n})$, where the family component assigns to
every natural number $n$ the two-point lattice $\downset{n} = \{n, \bot\}$ indicating whether $n$ is fully
present ($n$) or fully absent ($\bot$) from an approximation of $n$, as shown in the example in
\secref{introduction}. We could then interpret $+$ as the morphism $(+,\partial+)$, i.e.~the usual addition
operator + paired with an approximating version $\partial+_{n,m}: \downset{n} \to \downset{m}$ defined as:
\begin{displaymath}
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \partial+_{n,m}&n&m&=&n + m \\
      \partial+_{n,m}&\bot&\_&=&\bot \\
      \partial+_{n,m}&\_&\bot&=&\bot
    \end{array}
\end{displaymath}
These elementary approximation structures would then be combined by the interpretation of composite types like
sums and products, providing suitable derived structures for composite values.

@@

One might also like to allow the programmer to embed additional ``approximation points'' into composite
structures, by providing them with a \emph{lifting} type constructor $\synVar{L}$. Rather than the usual
definition of lists of natural numbers as $\mu\alpha.1 + \synVar{nat} \times \alpha$, one could then (within
the language) define a type of ``approximation lists'' $\mu\alpha.\synVar{L}\,(1 + \synVar{L}\,\synVar{nat}
\times \alpha)$, or perhaps a custom approximating version of multiplication $*_{\text{appr}}:
\synVar{L}\,\synVar{nat} \times \synVar{L}\,\synVar{nat} \to \synVar{L}\,\synVar{nat}$, expressible as
\[x *_{\text{appr}} y = \tmBind{x}{\tmFun{x'}{\syn{if}\;x' \equiv
0\;\syn{then}\;\tmReturn{0}\;\syn{else}\;\tmBind{y}{\tmFun{y'}{\tmReturn{x' * y'}}}}}\]

@@

>>>>

For a first-order language with a primitive type $\syn{nat}$ and primitive operations such as $+$, the
language would provide a model of $\syn{nat}$ and its operations in $\Fam(\LatGal)$. For example, we might
interpret $\syn{nat}$ to the object $(\Nat,n \mapsto \downset{n})$, where the family component assigns to
every natural number $n$ the two-point lattice $\downset{n} = \{n, \bot\}$ indicating whether $n$ is fully
present ($n$) or fully absent ($\bot$) from an approximation of $n$, as shown in the example in
\secref{introduction}. A slight variant is to use the constant family which assigns to every $n$ the two-point
lattice $\Two = \{\mathsf{tt},\mathsf{ff}\}$; using $\downset{n}$ is slightly more ergonomic as the
approximants of a given value then contain, rather than merely determining, all the information about the
value that hasn't been approximated away.

We might then interpret $+$ as the morphism $(+,\partial+)$, i.e.~the usual addition operator + paired with an
approximating version $\partial+_{n,m}: \downset{n} \to \downset{m}$ defined as:
\begin{displaymath}
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \partial+_{n,m}&n&m&=&n + m \\
      \partial+_{n,m}&\bot&\_&=&\bot \\
      \partial+_{n,m}&\_&\bot&=&\bot
    \end{array}
\end{displaymath}
These elementary approximation structures would then be combined by the interpretation of composite types like
sums and products, providing suitable derived structures for composite values.

@@

A programmer, or perhaps a language implementor, might also like to embed additional ``approximation points''
into composite structures, to allow for example the list $1 : 2 : []$ to have the partial list $1 : \bot$ as
an approximant, representing a list with at least one element. This can be accomplished by extending the
syntax with a \emph{lifting} type constructor $\synVar{L}$. Then, as an alternative to the usual definition of
lists of natural numbers as $\mu\alpha.1 + \synVar{nat} \times \alpha$, one could (within the language) define
a type of ``approximation lists'' $\mu\alpha.\synVar{L}\,(1 + \synVar{L}\,\synVar{nat} \times \alpha)$, with
monadic computations over approximated lists computing partial lists like $1 : \bot$. Similarly, one could
define a custom approximating version of multiplication $*_{\text{appr}}: \synVar{L}\,\synVar{nat} \times
\synVar{L}\,\synVar{nat} \to \synVar{L}\,\synVar{nat}$, as:
\[x *_{\text{appr}} y = \tmBind{x}{\tmFun{x'}{\syn{if}\;x' \equiv
0\;\syn{then}\;\tmReturn{0}\;\syn{else}\;\tmBind{y}{\tmFun{y'}{\tmReturn{x' * y'}}}}}\]

@@

With this lifting operator in the core language, one could then choose to fix certain choices of
approximation, by providing a surface language without the lifting monad and interpreting it into the core
language with lifting. The Galois slicing implementations discussed in \citet{perera12a} and
\citet{ricciotti17}, for example, provide an approximation point at every composite type constructor; we can
emulate this in our approach by defining a surface language where the data type of lists $\mu\alpha.1 +
\synVar{nat} \times \alpha$ would desugar automatically into the approximating version
$\mu\alpha.\synVar{L}\,(1 + \synVar{L}\,\synVar{nat} \times \alpha)$.
